\pkg{cubble}: An R Package for Structuring Spatio-temporal Data

### 2022-04-19 4-1-basic.Rmd
add example to cast covid data into cubble

>>>>

## Victoria covid data

This example illustrate casting different sources into a cubble object. Victoria government of Australia provides daily COVID case number by date, source, and local government area (LGA). This data can be useful for visualising COVID spread when combined with LGA map information, which is available from Australian Bureau of Statistics. Variable \code{covid} and \code{lga} show the first 5 rows of these two data:

```{r eval = FALSE}
library(tidyverse)
#https://www.coronavirus.vic.gov.au/victorian-coronavirus-covid-19-data
raw <- read_csv(here::here("data/NCOV_COVID_Cases_by_LGA_Source_20220324.csv"), 
                  col_names = c("date", "postcode", "source", "lga"), skip = 1)

covid <- raw %>%
  count(date, lga, source, postcode) %>%
  filter(date >= as.Date("2022-01-01")) %>% 
  filter(source == "Contact with a confirmed case") %>% 
  group_by(date, lga, source) %>%
  count() %>%
  ungroup(date) %>%
  mutate(roll_mean = zoo::rollmean(n, k = 14, fill = NA)) %>% 
  tsibble::as_tsibble(key = lga, index = date) %>% 
  tsibble::fill_gaps(.full = TRUE)
save(covid, file = here::here("data/covid.rda"))
```

```{r}
load(here::here("data/covid.rda"))
lga <- absmapsdata::lga2018 %>%
  rename(lga = lga_name_2018) %>% 
  filter(state_name_2016 == "Victoria") %>% 
  rmapshaper::ms_simplify()
```

```{r echo = TRUE}
covid |> head(5)
lga |> head(5)
```

To cast these two data into a cubble object, supply them as a list with name \code{spatial} and \code{temporal} and other arguments (\code{key}, \code{index}, and \code{coords}). \code{as_cubble()} will automatically check the match of spatial and temporal information in the two datasets and warnings will emit if a location has missing spatial or temporal information.

```{r echo = TRUE, warning = TRUE, message= TRUE}
cb <- as_cubble(list(spatial = lga, temporal = covid),
                key = lga, index = date, coords = c(cent_long, cent_lat))
```

Here, some locations have been detected with this issue and you can use \code{output = "unmatch"} to check on these locations:

```{r echo = TRUE}
pair <- as_cubble(list(spatial = lga, temporal = covid),
                key = lga, index = date, coords = c(cent_long, cent_lat),
                output = "unmatch")

pair
```

\pkg{cubble} will make an attempt to pair the unmatched sites as well as showing unpaired locations. These can be useful information to help clean up the data.

```{r echo = TRUE}
lga <- lga %>% 
  mutate(lga = ifelse(lga == "Kingston (C) (Vic.)", "Kingston (C)", lga),
         lga = ifelse(lga == "Latrobe (C) (Vic.)", "Latrobe (C)", lga)) %>%
  filter(!lga %in% pair$others$spatial)

covid <- covid %>% filter(!lga %in% pair$others$temporal)
```

After cleaning, you can cast the data into cubble with the same syntax without warnings:

```{r echo = TRUE}
cb <- as_cubble(data = list(spatial = lga, temporal = covid),
                key = lga, index = date, coords = c(cent_long, cent_lat))
```

### 2022-04-19 jss/cubble-jss.Rmd
notes on package install and rendering
a new skeleton of intro incorporating various past writing
add example to cast covid data into cubble

---
documentclass: jss
author:
  - name: H. Sherry Zhang
    affiliation: Monash University
    address: 21 Chancellors Walk, Clayton VIC 3800 Australia
    email: \email{huize.zhang@monash.edu}
  - name: Dianne Cook
    affiliation: 'Monash University \AND'
    address: 21 Chancellors Walk, Clayton VIC 3800 Australia
    # location: Melbourne, Australia
    email: dicook@monash.edu
  - name: Ursula Laa
    affiliation: 'University of Natural Resources and Life Sciences \AND'
    address: Gregor-Mendel-Straße 33, 1180 Wien, Austria
    email:  ursula.laa@boku.ac.at
  - name: Nicolas Langrené
    affiliation: 'BNU-HKBU United International College \AND'
    address: 2000 Jintong Road, Tangjiawan, Zhuhai, Guangdong Province, China
    email: nicolaslangrene@uic.edu.cn
  - name: Patricia Menéndez
    affiliation: 'Monash University \AND'
    address: 21 Chancellors Walk, Clayton VIC 3800 Australia
    email:  patricia.menendez@monash.edu
title:
  formatted: "\\pkg{cubble}: An R Package for Structuring Spatio-temporal Data"
  # If you use tex in the formatted title, alsoge supply version without
  plain:     "cubble: An R Package for Structuring Spatio-temporal Data"
  # For running headers, if needed
  # short:     "\\pkg{foo}: A Capitalized Title"
abstract: >
  Spatio-temporal data refer to measurements taken across space and time. In practice, spatio-temporal data can be decomposed into a spatial and temporal component: at one time, we would select a spatial location and inspect the temporal trend; at other time, we might select one or multiple time value(s) and explore the spatial distribution. Ideally, we could make multiple maps and multiple time series to explore these together, however, doing all of these actions is complicated when data arrive fragmented in multiple objects. To make it easy to do all these tasks, ideally spatial and temporal variables are in a single data object that we can slice and dice in different ways to conduct different visualisations. This work suggests a new data structure, \code{cubble}, that uses a nested form and a long form to organise spatial and temporal variables in a single data object. The new data structure ensures that data in the two forms are synchronised while been flexible to explore the two aspect of spatio-temporal data. It also provides capability for handling data with hierarchical structure, matching data from multiple sources, constructing interactive graphics, and performing spatio-temporal transformation. The paper will demonstrate \pkg{cubble} with examples using Australian climate weather stations, river level data, and climate reanalysis (ERA5) data.
keywords:
  # at least one keyword must be supplied
  formatted: [spatial, temporal, spatio temporal, "\\proglang{R}", environmental data, exploratory data analysis]
  plain:     [spatial, temporal, spatio temporal, R, environmental data, exploratory data analysis]
preamble: >
  \usepackage{amsmath}
  \usepackage{array}
  \usepackage{xcolor}
output: rticles::jss_article
bibliography: references.bib
---

```{r setup, echo = FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  cache = TRUE,
  fig.path = "figures/",
  fig.height = 10
)
options(prompt = "R> ", continue = "+ ",
        tibble.print_max = 5, tibble.print_min = 5)
```

>>>>

---
documentclass: jss
author:
  - name: H. Sherry Zhang
    affiliation: Monash University
    address: 21 Chancellors Walk, Clayton VIC 3800 Australia
    email: \email{huize.zhang@monash.edu}
  - name: Dianne Cook
    affiliation: 'Monash University \AND'
    address: 21 Chancellors Walk, Clayton VIC 3800 Australia
    # location: Melbourne, Australia
    email: dicook@monash.edu
  - name: Ursula Laa
    affiliation: 'University of Natural Resources and Life Sciences \AND'
    address: Gregor-Mendel-Straße 33, 1180 Wien, Austria
    email:  ursula.laa@boku.ac.at
  - name: Nicolas Langrené
    affiliation: 'BNU-HKBU United International College \AND'
    address: 2000 Jintong Road, Tangjiawan, Zhuhai, Guangdong Province, China
    email: nicolaslangrene@uic.edu.cn
  - name: Patricia Menéndez
    affiliation: 'Monash University \AND'
    address: 21 Chancellors Walk, Clayton VIC 3800 Australia
    email:  patricia.menendez@monash.edu
title:
  formatted: "\\pkg{cubble}: An R Package for Structuring Spatio-temporal Data"
  # If you use tex in the formatted title, alsoge supply version without
  plain:     "cubble: An R Package for Structuring Spatio-temporal Data"
  # For running headers, if needed
  # short:     "\\pkg{foo}: A Capitalized Title"
abstract: >
  Spatio-temporal data refer to measurements taken across space and time. In practice, spatio-temporal data can be decomposed into a spatial and temporal component: at one time, we would select a spatial location and inspect the temporal trend; at other time, we might select one or multiple time value(s) and explore the spatial distribution. Ideally, we could make multiple maps and multiple time series to explore these together, however, doing all of these actions is complicated when data arrive fragmented in multiple objects. To make it easy to do all these tasks, ideally spatial and temporal variables are in a single data object that we can slice and dice in different ways to conduct different visualisations. This work suggests a new data structure, \code{cubble}, that uses a nested form and a long form to organise spatial and temporal variables in a single data object. The new data structure ensures that data in the two forms are synchronised while been flexible to explore the two aspect of spatio-temporal data. It also provides capability for handling data with hierarchical structure, matching data from multiple sources, constructing interactive graphics, and performing spatio-temporal transformation. The paper will demonstrate \pkg{cubble} with examples using Australian climate weather stations, river level data, and climate reanalysis (ERA5) data.
keywords:
  # at least one keyword must be supplied
  formatted: [spatial, temporal, spatio temporal, "\\proglang{R}", environmental data, exploratory data analysis]
  plain:     [spatial, temporal, spatio temporal, R, environmental data, exploratory data analysis]
preamble: >
  \usepackage{amsmath}
  \usepackage{array}
  \usepackage{xcolor}
output: 
  bookdown::pdf_book:
    base_format: rticles::jss_article
bibliography: ../references.bib
---

```{r setup, echo = FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  cache = TRUE,
  fig.path = "figures/",
  fig.height = 10
)

options(prompt = "R> ", continue = "+ ",
        tibble.print_max = 5, tibble.print_min = 5)
```

### 2022-04-19 jss/cubble-jss.tex
add example to cast covid data into cubble

Spatio-temporal data has a spatial component referring to the location of each observation, a temporal component that is recorded at regular or irregular time intervals, and multiple variables measured at each spatial and temporal value. In data collection, spatio-temporal data is less often collected as a whole piece but fragmented spatial and temporal data for analysts to ensemble into spatio-temporal data. With the {[}integrated{]} data, we hope to fix the time to explore the spatial distribution, or fix the spaital location to explore temporal trends. Ideally, we would be able to carry out each of these tasks easily, and even accommodate complex explorations encompassing both simultaneously. These are all underpinned by a {[}flexible{]} data structure that can represent spatio-temporal data.

@@

What remains to be difficult when wrangling spatio-temporal data is that the raw data is not naturally structured in any one of the representation mentioned above and the first difficulty is to cast the data, potentially from multiple sources into one of the data representation. Some examples of these includes map data recorded the shape of a collection of area of interest is provided by the government website; geo-scientific data recorded with longitude and latitude coordinate of locations in the area; and temporal variables indexed by both location and time for the temporal aspect. For spatio-temporal data analysis, merging all these sources can be challenging and joining these tables can go wrong because of the slightest unmatched of the linking variable from different sources.

Another issue is how the concept of tidy data should be applied to spatio-temporal data. The tidy data concept advocates the data to be shaped into 1) one row per observation; 2) one column per variable; and 3) one type of data per table. The long form data is preferred over wide data given the downstream data cleaning and visualisation software i.e.~\pkg{dplyr} and \pkg{ggplot2}. However, the long form can be inefficient to store feature geometries, especially for large multipolygon over long time period, i.e.~sub-daily time series data.

@@

@@

>>>>

Spatio-temporal data has a spatial component referring to the location of each observation, a temporal component that is recorded at regular or irregular time intervals, and multiple variables measured at each spatial and temporal value. The analysis of spatio-temporal data can fix the time to explore the spatial distribution, fix the spatial location to explore temporal trends, or dynamically explore the space and time simultaneously. For these tasks to be carried out easily, there is a need for a spatio-temporal data representation that underpins various operations.

@@

What remains to be difficult when wrangling spatio-temporal data is that the raw data is not naturally structured in any one of the representation above and the first task is to cast the data into a class. Also, in data collection, spatio-temporal data is less often collected as a whole piece but fragmented spatial and temporal data for analysts to ensemble. For example, map data records the shape of a collection of area of interest is provided by official government website; geo-scientific data records the longitude and latitude coordinate of locations in the area along with other meta data relates to the location; and temporal variables indexed by both location and time. For spatio-temporal data analysis, merging all these sources can be challenging and joining these tables can go wrong because of the slightest unmatched of the linking variable from different sources.

Another issue is how the concept of tidy data \citep{tidydata} should be applied to spatio-temporal data. The tidy data concept advocates the data to be shaped into 1) one row per observation; 2) one column per variable; and 3) one type of data per table. The long form data is preferred over wide data given the downstream software i.e.~\pkg{dplyr} and \pkg{ggplot2} for data wrangling and visualisation. However, the long form can be inefficient to store feature geometries, especially for large multipolygon over long time period. This poses the question on how to arrange spatial and temporal variables that makes analysing spatio-temporal data easier.

@@

\hypertarget{why-not-just-use-dplyrnestby}{%
\subsection{Why not just use dplyrnestby}\label{why-not-just-use-dplyrnestby}}

@@

\hypertarget{victoria-covid-data}{%
\subsection{Victoria covid data}\label{victoria-covid-data}}

This example illustrate casting different sources into a cubble object. Victoria government of Australia provides daily COVID case number by date, source, and local government area (LGA). This data can be useful for visualising COVID spread when combined with LGA map information, which is available from Australian Bureau of Statistics. Variable \code{covid} and \code{lga} show the first 5 rows of these two data:

\begin{CodeChunk}
\begin{CodeInput}
R> covid |> head(5)
\end{CodeInput}
\begin{CodeOutput}
# A tsibble: 5 x 5 [1D]
# Key:       lga [1]
# Groups:    lga, source [1]
  date       lga        source                            n roll_mean
  <date>     <chr>      <chr>                         <int>     <dbl>
1 2022-01-01 Alpine (S) Contact with a confirmed case     1        NA
2 2022-01-02 Alpine (S) Contact with a confirmed case     2        NA
3 2022-01-03 Alpine (S) Contact with a confirmed case     4        NA
4 2022-01-04 Alpine (S) Contact with a confirmed case     4        NA
5 2022-01-05 Alpine (S) Contact with a confirmed case     2        NA
\end{CodeOutput}
\begin{CodeInput}
R> lga |> head(5)
\end{CodeInput}
\begin{CodeOutput}
Simple feature collection with 5 features and 7 fields
Geometry type: MULTIPOLYGON
Dimension:     XY
Bounding box:  xmin: 142.3643 ymin: -38.67876 xmax: 147.386 ymax: -36.39269
Geodetic CRS:  WGS 84
  lga_code_2018            lga state_code_2016 state_name_2016 areasqkm_2018
1         20110     Alpine (S)               2        Victoria     4788.1568
2         20260    Ararat (RC)               2        Victoria     4211.1171
3         20570   Ballarat (C)               2        Victoria      739.0321
4         20660    Banyule (C)               2        Victoria       62.5402
5         20740 Bass Coast (S)               2        Victoria      865.8095
  cent_long  cent_lat                       geometry
1  146.9742 -36.85357 MULTIPOLYGON (((146.7258 -3...
2  142.8432 -37.47271 MULTIPOLYGON (((143.1807 -3...
3  143.7815 -37.49286 MULTIPOLYGON (((143.6622 -3...
4  145.0851 -37.73043 MULTIPOLYGON (((145.1357 -3...
5  145.5581 -38.50730 MULTIPOLYGON (((145.5207 -3...
\end{CodeOutput}
\end{CodeChunk}

To cast these two data into a cubble object, supply them as a list with name \code{spatial} and \code{temporal} and other arguments (\code{key}, \code{index}, and \code{coords}). \code{as_cubble()} will automatically check the match of spatial and temporal information in the two datasets and warnings will emit if a location has missing spatial or temporal information.

\begin{CodeChunk}
\begin{CodeInput}
R> cb <- as_cubble(list(spatial = lga, temporal = covid),
+                 key = lga, index = date, coords = c(cent_long, cent_lat))
\end{CodeInput}
\begin{CodeOutput}
! Some sites in the temporal table don't have corresponding spatial information
\end{CodeOutput}
\begin{CodeOutput}
! Some sites in the spatial table don't have corresponding temporal information
\end{CodeOutput}
\begin{CodeOutput}
! Use argument `output = "unmatch"` to check on the unmatched key
\end{CodeOutput}
\end{CodeChunk}

Here, some locations have been detected with this issue and you can use \code{output = "unmatch"} to check on these locations:

\begin{CodeChunk}
\begin{CodeInput}
R> pair <- as_cubble(list(spatial = lga, temporal = covid),
+                 key = lga, index = date, coords = c(cent_long, cent_lat),
+                 output = "unmatch")
R> 
R> pair
\end{CodeInput}
\begin{CodeOutput}
$paired
# A tibble: 2 x 2
  spatial             temporal    
  <chr>               <chr>       
1 Kingston (C) (Vic.) Kingston (C)
2 Latrobe (C) (Vic.)  Latrobe (C)

$others
$others$temporal
[1] "Interstate"       "Overseas"         "Queenscliffe (B)" "Unknown"

$others$spatial
character(0)
\end{CodeOutput}
\end{CodeChunk}

\pkg{cubble} will make an attempt to pair the unmatched sites as well as showing unpaired locations. These can be useful information to help clean up the data.

\begin{CodeChunk}
\begin{CodeInput}
R> lga <- lga %>% 
+   mutate(lga = ifelse(lga == "Kingston (C) (Vic.)", "Kingston (C)", lga),
+          lga = ifelse(lga == "Latrobe (C) (Vic.)", "Latrobe (C)", lga)) %>%
+   filter(!lga %in% pair$others$spatial)
R> 
R> covid <- covid %>% filter(!lga %in% pair$others$temporal)
\end{CodeInput}
\end{CodeChunk}

After cleaning, you can cast the data into cubble with the same syntax without warnings:

\begin{CodeChunk}
\begin{CodeInput}
R> cb <- as_cubble(data = list(spatial = lga, temporal = covid),
+                 key = lga, index = date, coords = c(cent_long, cent_lat))
\end{CodeInput}
\end{CodeChunk}
